Predicting instabilities of a tuneable ring laser with an iterative map model

### 2021-02-21 Papers/Journal/Tuneable_Lasers.tex
Fixed formatting

\usepackage{subcaption}
\usepackage{tikz}
\usepackage{mathtools}

@@

\section{Introduction}
\label{sec:intro}
Dispersion-tuned actively mode-locked lasers are an active area of optics research since they have a wide range of applications. These lasers can have ultrashort pulses, as short as 50 femtoseconds \cite{chung2017}, and the ability to tune their output frequency rapidly over a range of over 100 nm~\cite{bohun2015, burgoyne2010, chung2017, yamashita2009}. The capability to alter the operating frequency makes tuneable lasers particularly useful for imaging applications, such as optical coherence tomography~\cite{bohun2015, burgoyne2014, yamashita2009}, coherent anti-Stokes Raman spectroscopy~\cite{burgoyne2014}, and deep tissue multi-photon microscopy~\cite{chung2017}, as well as sensing and measuring of other ultrafast processes~\cite{burgoyne2014, silfvast2004, oktem2010}. Tuneable lasers are usually constructed in a ring with five main components: an optical coupler, a chirped fibre Bragg grating (CFBG), a modulator (or saturable absorber), a gain fibre (usually doped with Er), and a pump laser. A typical ring laser cavity is depicted in Fig.~\ref{fig:cavity}. The Kerr nonlinearity plays a crucial role in the dynamics within the cavity due to the high power and ultrashort duration of the pulses.

@@

To resolve some of the issues with an `average' model, each component and region in the laser is modelled with a separate GNLSE. The solution of one GNLSE is used as the initial condition for the GNLSE for the next component~\cite{lapre2019, meng2020, oktem2010, woodward2018}. An iterative map model allows for better encapsulation of specific dynamics introduced by each component. The main drawback of this method is the extreme computational expense since a nonlinear partial differential equation must be numerically solved for every segment for each of the hundreds of round trips in a given simulation. We can alleviate this computational expense by noticing each effect is localized to a corresponding component in the cavity: a majority of the loss is due to the optical coupler, virtually all of the dispersion happens within the CFBG~\cite{agrawal2002}, the pulse is only modulated by the modulator, and the pulse is amplified only within the Er-doped gain fibre. Therefore, we can drastically simplify each of the GNLSEs into a form we can solve analytically. Using analytic maps instead of numerically solving a GNLSE reduces the computational cost, while maintaining the benefit of reflecting the specific geometry of a component-wise simulation.

Iterative map methods have been used for optics since 1955 when Cutler~\cite{cutler1955} analyzed a microwave regenerative pulse generator. Siegman and Kuizenga~\cite{kuizenga1970a, kuizenga1970b, kuizenga1970, siegman1969} adapted Cutler's method for both AM and FM mode-locked lasers in 1969, and Martinez \emph{et al.}~\cite{martinez1984, martinez1985} incorporated the nonlinearity while modelling passively mode-locked lasers. A linear iterative method was used for dispersion-tuned actively mode-locked lasers recently by Burgoyne \emph{et al.}~\cite{burgoyne2014}. Each of these models used a prescribed transfer function to describe the effect each component has on the pulse. While the transfer functions are chosen suitably to have the effect observed in the laboratory, the transfer functions are phenomenologically chosen, and not necessarily obtained directly from the underlying physics. Furthermore, these methods have not included the effect of every component, lacking the imperative interactions of dispersion, modulation, and nonlinearity---hindering their models to exhibit modulation instability.

@@

\subsection{Combining the Component Maps}
\label{sec:effects}

@@

Now that we have the algebraic effect of each component of the cavity, we are ready to take the composition of the maps to give the effect of one round trip of the cavity. Unlike the models described in Section \ref{sec:intro}, the order of our components is important since the maps in \eqref{eq:dimgain}, and \eqref{eq:dimnl}--\eqref{eq:modform} do not necessarily commute with each other.

@@

% Table of parameter values
\begin{table*}[tbp]
	\centering
	\caption{Range of parameter values (taken from~\cite{agrawal2013, burgoyne2014, burgoyneemail, tamura1996, usechak2005}).}
 	\label{tab:values}
 	\begin{tabular}{clll}
		\hline
		Parameter & Description & Value & Units \\
		\hline\noalign{\smallskip}
		$T_M$ & Characteristic modulation time & $15$--$150$ & ps \\
		$\beta_2 L_D$ & Dispersion provided by the CFBG & $10$--$2000$ & ps$^2$ \\
		$\gamma$ & Kerr nonlinearity coefficient & $0.001$--$0.01$ & W$^{-1}$ m$^{-1}$ \\
		$L_f$ & Length of fibre between gain and optical coupler & $0.15$--$1$ & m \\
		$L_g$ & Length of the Er-doped gain fibre & $2$--$3$ & m \\
		$\alpha$ & Loss of fibre due to scattering and absorption & $10^{-4}$--$0.3$ & m$^{-1}$ \\
		$R$ & Reflectivity of optical coupler & $0.1$--$0.9$ & -- \\
		$\Es$ & Saturation energy of Er-doped gain fibre & $10^3$--$10^4$ & pJ \\
		$g_0$ & Small signal gain &  $1$--$10$ & m$^{-1}$ \\
		$L_T$ & Total length of fibre in the cavity & $10$--$100$ & m \\
		\noalign{\smallskip}\hline
	\end{tabular}
\end{table*}

@@

We consider the pulse given by
\begin{align}
	A = \sqrt{P} \exp \left( -(1 + iC) \frac{T^2}{2 \sigma^2} \right) \exp(i \phi_0),
	\label{eq:A0}
\end{align}
where $P$ is the peak power, $C$ is the chirp, $\sigma^2$ is the variance, and $\phi_0$ is the phase; the pulse is shown in Fig.~\ref{fig:samplegauss}. By computing \eqref{eq:A0} after one round trip, and solving $\mathcal{L}(A) = A \textrm{e}^{i \phi}$, tedious manipulations yield the conditions
\begin{align}
	\frac{\sigma^4}{1 - \sigma^2} &= \left( \sigma^2 + 2 C s^2 \right)^2 + 4 s^4, \label{eq:sigrelation} \\
	\frac{C}{1 - \sigma^2} &= C + 2 \frac{s^2}{\sigma^2} \left( 1 + C^2 \right), \label{eq:chirprelation} \\
	1 &= \frac{W(a E_g \textrm{e}^{E_g})}{E_g} h^2 \left( 1 - \sigma^2 \right)^{1/2} \label{eq:energyrelation}
\end{align}
for \eqref{eq:A0} to be in equilibrium. Using \eqref{eq:sigrelation} and \eqref{eq:chirprelation} we eliminate the chirp, $C$, and find an expression relating $\sigma^2$ and $s$, namely
\begin{align}
	\sigma^8 + 4 s^4 \sigma^6 - 20 s^4 \sigma^4 + 32 s^4 \sigma^2 - 16 s^4 = 0.
	\label{eq:var}
\end{align}
\eqref{eq:var} has the positive, real, analytic solution
\begin{equation}
	\begin{split}
		\sigma^2 = \sqrt{2} s \left( s^6 + 3s^2 + \sqrt{4 + s^4} \left( 1 + s^4 \right) \right)^{1/2} & \\
		- s^4 - s^2 & \sqrt{4 + s^4}.
		\label{eq:equilvar}
	\end{split}
\end{equation}
We find the chirp is
\begin{align}
	C = \frac{\sigma^4 - \left( \sigma^8 - 16 s^4 \left( 1 - \sigma^2 \right)^2 \right)^{1/2}}{4 s^2 \left( 1 - \sigma^2 \right)}
	\label{eq:chirp}
\end{align}
by \eqref{eq:chirprelation}. We obtain the energy of the pulse entering the gain fibre, $E_g$, from \eqref{eq:energyrelation}:
\begin{align}
	E_g &= \frac{h^2 \left( 1 - \sigma^2 \right)^{1/2} \log \left( a h^2 \left( 1 - \sigma^2 \right)^{1/2} \right)}{h^2 \left( 1 - \sigma^2 \right)^{1/2} - 1},
\end{align}
and thus, the equilibrium energy at the optical coupler, is
\begin{align}
	E &= W \left( a E_g \textrm{e}^{E_g} \right), \\
	&= \frac{\log \left( a h^2 \left( 1 - \sigma^2 \right)^{1/2} \right)}{1 - h^2 \left( 1 - \sigma^2 \right)^{1/2}}.
	\label{eq:analenergy}
\end{align}
The argument of the logarithm in \eqref{eq:analenergy} gives us the condition
\begin{align}
	a h^2 (1 - \sigma^2)^{1/2} > 1
	\label{eq:energycond}
\end{align}
for the pulse to be sustainable.

>>>>

\usepackage{subcaption}
\usepackage{tikz}
\usepackage{mathtools}
\usepackage{dblfloatfix}

@@

\section{Introduction}
\label{sec:intro}
Dispersion-tuned actively mode-locked lasers are an active area of optics research since they have a wide range of applications. These lasers can have ultrashort pulses, as short as 50 femtoseconds \cite{chung2017}, and the ability to tune their output frequency rapidly over a range of over 100 nm~\cite{bohun2015, burgoyne2010, chung2017, yamashita2009}. The capability to alter the operating frequency makes tuneable lasers particularly useful for imaging applications, such as optical coherence tomography~\cite{bohun2015, burgoyne2014, yamashita2009}, coherent anti-Stokes Raman spectroscopy~\cite{burgoyne2014}, and deep tissue multi-photon microscopy~\cite{chung2017}, as well as sensing and measuring of other ultrafast processes~\cite{burgoyne2014, silfvast2004, oktem2010}. Tuneable lasers are usually constructed in a ring with five main components: an optical coupler, a chirped fibre Bragg grating (CFBG), a modulator (or saturable absorber), a gain fibre (usually doped with Er), and a pump laser. A typical ring laser cavity is depicted in Fig.~\ref{fig:cavity}. The Kerr nonlinearity plays a crucial role in the dynamics within the cavity due to the high power and ultrashort duration of the~pulses.

@@

To resolve some of the issues with an `average' model, each component and region in the laser is modelled with a separate GNLSE. The solution of one GNLSE is used as the initial condition for the GNLSE for the next component~\cite{lapre2019, meng2020, oktem2010, woodward2018}. An iterative map model allows for better encapsulation of specific dynamics introduced by each component. The main drawback of this method is the extreme computational expense since a nonlinear partial differential equation must be numerically solved for every segment for each of the hundreds of round trips in a given simulation. We can alleviate this computational expense by noticing each effect is localized to a corresponding component in the cavity: a majority of the loss is due to the optical coupler, virtually all of the dispersion happens within the CFBG~\cite{agrawal2002}, the pulse is only modulated by the modulator, and the pulse is amplified only within the Er-doped gain fibre. Therefore, we can drastically simplify each of the GNLSEs into a form we can solve analytically. Using analytic maps instead of numerically solving a GNLSE reduces the computational cost, while maintaining the benefit of reflecting the specific geometry of a component-wise~simulation.

Iterative map methods have been used for optics since 1955 when Cutler~\cite{cutler1955} analyzed a microwave regenerative pulse generator. Siegman and Kuizenga~\cite{kuizenga1970a, kuizenga1970b, kuizenga1970, siegman1969} adapted Cutler's method for both AM and FM mode-locked lasers in 1969, and Martinez \emph{et al.}~\cite{martinez1984, martinez1985} incorporated the nonlinearity while modelling passively mode-locked lasers. A linear iterative method was used for dispersion-tuned actively mode-locked lasers recently by Burgoyne \emph{et al.}~\cite{burgoyne2014}. Each of these models used a prescribed transfer function to describe the effect each component has on the pulse. While the transfer functions are chosen suitably to have the effect observed in the laboratory, the transfer functions are phenomenologically chosen, and not necessarily obtained directly from the underlying physics. Furthermore, these methods have not included the effect of every component, lacking the interactions of dispersion, modulation, and nonlinearity---hindering their models ability to exhibit modulation instability.

@@

\subsection{Combining the Component Maps}
\label{sec:effects}
Now that we have the algebraic effect of each component of the cavity, we are ready to take the composition of the maps to give the effect of one round trip of the cavity. Unlike the models described in Section \ref{sec:intro}, the order of our components is important since the maps in \eqref{eq:dimgain}, and \eqref{eq:dimnl}--\eqref{eq:modform} do not necessarily commute with each other.

@@

% Table of parameter values
\begin{table*}[!b]
	\centering
	\caption{Range of parameter values (taken from~\cite{agrawal2013, burgoyne2014, burgoyneemail, tamura1996, usechak2005}).}
 	\label{tab:values}
 	\begin{tabular}{clll}
		\hline
		Parameter & Description & Value & Units \\
		\hline\noalign{\smallskip}
		$T_M$ & Characteristic modulation time & $15$--$150$ & ps \\
		$\beta_2 L_D$ & Dispersion provided by the CFBG & $10$--$2000$ & ps$^2$ \\
		$\gamma$ & Kerr nonlinearity coefficient & $0.001$--$0.01$ & W$^{-1}$ m$^{-1}$ \\
		$L_f$ & Length of fibre between gain and optical coupler & $0.15$--$1$ & m \\
		$L_g$ & Length of the Er-doped gain fibre & $2$--$3$ & m \\
		$\alpha$ & Loss of fibre due to scattering and absorption & $10^{-4}$--$0.3$ & m$^{-1}$ \\
		$R$ & Reflectivity of optical coupler & $0.1$--$0.9$ & -- \\
		$\Es$ & Saturation energy of Er-doped gain fibre & $10^3$--$10^4$ & pJ \\
		$g_0$ & Small signal gain &  $1$--$10$ & m$^{-1}$ \\
		$L_T$ & Total length of fibre in the cavity & $10$--$100$ & m \\
		\noalign{\smallskip}\hline
	\end{tabular}
\end{table*}

@@

@@

We consider the pulse given by
\begin{align}
	A = \sqrt{P} \exp \left( -(1 + iC) \frac{T^2}{2 \sigma^2} \right) \exp(i \phi_0),
	\label{eq:A0}
\end{align}
where $P$ is the peak power, $C$ is the chirp, $\sigma^2$ is the variance, and $\phi_0$ is the phase; the pulse is shown in Fig.~\ref{fig:samplegauss}. By computing \eqref{eq:A0} after one round trip, and solving $\mathcal{L}(A) = A \textrm{e}^{i \phi}$, tedious manipulations yield the conditions
\begin{align}
	\frac{\sigma^4}{1 - \sigma^2} &= \left( \sigma^2 + 2 C s^2 \right)^2 + 4 s^4, \label{eq:sigrelation} \\
	\frac{C}{1 - \sigma^2} &= C + 2 \frac{s^2}{\sigma^2} \left( 1 + C^2 \right), \label{eq:chirprelation} \\
	1 &= \frac{W(a E_g \textrm{e}^{E_g})}{E_g} h^2 \left( 1 - \sigma^2 \right)^{1/2} \label{eq:energyrelation}
\end{align}
for \eqref{eq:A0} to be in equilibrium. Using \eqref{eq:sigrelation} and \eqref{eq:chirprelation} we eliminate the chirp, $C$, and find an expression relating $\sigma^2$ and $s$,~namely
\begin{align}
	\sigma^8 + 4 s^4 \sigma^6 - 20 s^4 \sigma^4 + 32 s^4 \sigma^2 - 16 s^4 = 0.
	\label{eq:var}
\end{align}
\eqref{eq:var} has the positive, real, analytic solution
\begin{equation}
	\begin{split}
		\sigma^2 = \sqrt{2} s \left( s^6 + 3s^2 + \sqrt{4 + s^4} \left( 1 + s^4 \right) \right)^{1/2} & \\
		- s^4 - s^2 & \sqrt{4 + s^4}.
		\label{eq:equilvar}
	\end{split}
\end{equation}
We find the chirp is
\begin{align}
	C = \frac{\sigma^4 - \left( \sigma^8 - 16 s^4 \left( 1 - \sigma^2 \right)^2 \right)^{1/2}}{4 s^2 \left( 1 - \sigma^2 \right)}
	\label{eq:chirp}
\end{align}
by \eqref{eq:chirprelation}. We obtain the energy of the pulse entering the gain fibre, $E_g$, from \eqref{eq:energyrelation}:
\begin{align}
	E_g &= \frac{h^2 \left( 1 - \sigma^2 \right)^{1/2} \log \left( a h^2 \left( 1 - \sigma^2 \right)^{1/2} \right)}{h^2 \left( 1 - \sigma^2 \right)^{1/2} - 1},
\end{align}
and thus, the equilibrium energy at the optical coupler, is
\begin{align}
	E &= W \left( a E_g \textrm{e}^{E_g} \right), \\
	&= \frac{\log \left( a h^2 \left( 1 - \sigma^2 \right)^{1/2} \right)}{1 - h^2 \left( 1 - \sigma^2 \right)^{1/2}}.
	\label{eq:analenergy}
\end{align}
The argument of the logarithm in \eqref{eq:analenergy} gives us the condition
\begin{align}
	a h^2 (1 - \sigma^2)^{1/2} > 1
	\label{eq:energycond}
\end{align}
for the pulse to be sustainable.
